On mapping errors and seeding heuristics for DNA alignment

### 2019-04-09 MEM.tex
fixed typos sections 1-4.

The index is usually a hash table or a variant of the so-called
FM-index~\cite{pmid20460430}. Hash tables are typically used to index
$k$-mers, which makes them useful to search seeds of set size $k$,
(see~\cite{pmid30346548} for a recent benchmark of $k$-mer hashing
algorithms). In contrast, the FM-index has no set size so it can be used
to search seeds of different lengths~\cite{ferragina2000opportunistic,
ferragina2005indexing}. It is a compact data structure based on the suffix
array~\cite{manber1993suffix} and on the Burrows-Wheeler
transform~\cite{burrows1994block}, emulating a suffix trie with a much
smaller memory footprint.

@@

\begin{definition}
\label{def:seg}
Every symbol of the alphabet that is different from $\square$ is called a
terminator. A segment is a sequence of 0 or more symbols $\square$
followed by a terminator. The last segment of the read is terminated by
the symbol $|$ and is called the tail.
\end{definition}

@@

The weighted generating function of interest is the top right entry of the
matrix $M_n(z) + M_n(z)^2 + \ldots = M_n(z)\cdot(I-M_n(z))^{-1}$. To see
why, observe that at the start of every read, the next nucleotide is a
non-skipped position, so every read can be prepended by $\Downarrow_0$
segments and only by those. Thus, reads are precisely the sequences of
segments that can follow the symbol $\Downarrow_0$ and that are terminated
by a tail, whose weighted generating function is the entry of the matrix
associated with terminators $\Downarrow_0$ and $|$.

$M_n(z)$ is too complex to compute a closed expression of $M_n(z) \cdot
(I-M_n(z))^{-1}$. We return to the definition $M_n(z) + M_n(z)^2 + \ldots$
and observe that the terms $M_n(z)^{k+2}, M_n(z)^{k+3}, \ldots$ have no
influence on the coefficients of the weighted generating function $a_0,
a_1, \ldots, a_k$.

Indeed, a read of size $k$ has at most $k+1$ segments (including the
obligatory tail). Since $M(z)^{s+1}$ contains the weighted generating
functions of reads with exactly $s+1$ segments including the tail, $a_k$
cannot depend on $M(z)^{k+2}, M(z)^{k+3}$. More formally, we can prove by
induction that all the entries of $M(z)^k$ are divisible by $z^{k-1}$,
showing that the contribution of $M(z)^{k+2} + M(z)^{k+3} + \ldots$ to
$a_0 + a_1z + \ldots +a_kz^k$ is strictly $0$.

@@

The number $i$ in the symbol $\downarrow_{/i}^-$ indicates the match
length of the $+$ sequence. Likewise, the number $i$ in the symbol
$\downarrow_{/i}^+$ indicates the match length of the $-$ sequence. For
instance the symbol $\downarrow_{/7}^-$ indicates that the nucleotide is a
mismatch against the $-$ sequence, that it is a match for the $+$
sequence, and that the six previous nucleotides were also a match for the
$+$ sequence. The terminators thus encode the local state of the read.

>>>>

The index is usually a hash table or a variant of the so-called
FM-index~\cite{pmid20460430}. Hash tables are typically used to index
$k$-mers, which makes them useful to search seeds of set size $k$
(see~\cite{pmid30346548} for a recent benchmark of $k$-mer hashing
algorithms). In contrast, the FM-index has no set size so it can be used
to search seeds of different lengths~\cite{ferragina2000opportunistic,
ferragina2005indexing}. It is a compact data structure based on the suffix
array~\cite{manber1993suffix} and on the Burrows-Wheeler
transform~\cite{burrows1994block}, emulating a suffix trie with a much
smaller memory footprint.

@@

\begin{definition}
\label{def:seg}
Every symbol of the alphabet that is different from $\square$ is called a
terminator. A segment is a terminator preceded by 0 or more symbols $\square$.
The last segment of the read is terminated by the symbol $|$ and is called the
tail.
\end{definition}

@@

The weighted generating function of interest is the top right entry of the
matrix $M_n(z) + M_n(z)^2 + \ldots = M_n(z)\cdot(I-M_n(z))^{-1}$. To see
why, observe that, at the start of every read, the next nucleotide is a
non-skipped position, so every read can be prepended by $\Downarrow_0$
segments and only by those. Thus, reads are precisely the sequences of
segments that can follow the symbol $\Downarrow_0$ and that are terminated
by a tail, whose weighted generating function is the entry of the matrix
associated with terminators $\Downarrow_0$ and $|$.

$M_n(z)$ is too complex to allow computing a closed expression of $M_n(z) \cdot
(I-M_n(z))^{-1}$. We return to the definition $M_n(z) + M_n(z)^2 + \ldots$ and
observe that the terms $M_n(z)^{k+2}, M_n(z)^{k+3}, \ldots$ have no influence on
the coefficients of the weighted generating function $a_0, a_1, \ldots, a_k$.

Indeed, a read of size $k$ has at most $k+1$ segments (including the obligatory
tail). Since $M(z)^{s+1}$ contains the weighted generating functions of reads
with exactly $s+1$ segments including the tail, $a_k$ cannot depend on
$M(z)^{k+2}, M(z)^{k+3}, \ldots$. More formally, we can prove by induction that
all the entries of $M(z)^k$ are divisible by $z^{k-1}$, showing that the
contribution of $M(z)^{k+2} + M(z)^{k+3} + \ldots$ to $a_0 + a_1z + \ldots
+a_kz^k$ is strictly $0$.

@@

The number $i$ in the symbol $\downarrow_{/i}^-$ indicates the match
length of the $+$ sequence. Likewise, the number $i$ in the symbol
$\downarrow_{/i}^+$ indicates the match length of the $-$ sequence. For
instance, the symbol $\downarrow_{/7}^-$ indicates that the nucleotide is a
mismatch against the $-$ sequence, that it is a match for the $+$
sequence, and that the six previous nucleotides were also a match for the
$+$ sequence. The terminators thus encode the local state of the read.
